Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-04-01 experiments.tex
Finetuning

\begin{figure}[h]
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
\caption{Finding generating models of 7 arcs.}
\label{fig:4arcs}
\end{figure}

Not surprisingly, the results indicate that when the parameter generation
mechanism matches the assumptions of the BDeu-score, BDeu usually
also ranks the generating structure higher than the other scores
(Figure X).  On the other hand, when the parameters are drawn from the
Jeffreys' prior, the fNML score appears to rank the generating structure
highest. This too is not surprising, since the NML distribution for
multinomial data is known to closely match the marginal distribution of
the data when the Jeffreys' prior is used.

The underfitting tendency of BIC can also be clearly detected both for
relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
ranking ability the qNML criterion appears to be between fNML and BDeu
in 3 out of 4 settings which is good considering that one of these
criteria is always the winner. For the dense networks with BDeu prior,
qNML appears to perform worse than BDeu and fNML but still much
better than BIC.

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
\caption{Finding generating models of 7 arcs.}
\label{fig:7arcs}
\end{figure}

Not surprisingly, the results indicate that when the parameter generation
mechanism matches the assumptions of the BDeu-score, BDeu usually
also ranks the generating structure higher than the other scores (see Figure
~\ref{fig:4arcs} and \ref{fig:7arcs}).  On the other hand, when the parameters are drawn from the
Jeffreys' prior, the fNML score appears to rank the generating structure
highest. This too is not surprising, since the NML distribution for
multinomial data is known to closely match the marginal distribution of
the data when the Jeffreys' prior is used.

The underfitting tendency of BIC can also be clearly detected both for
relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
ranking ability of the qNML criterion appears to be between fNML and BDeu
in 3 out of 4 settings which is good considering that one of these
criteria is always the winner. For dense networks with the BDeu prior,
qNML appears to perform worse than BDeu and fNML but still much
better than BIC.

### 2017-04-01 intro.tex
Finetuning

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion (AIC)~\cite{Akai73} and the Bayesian information
criterion (BIC)~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for AIC, and BIC appears to
require large sample sizes in order to identify appropriate
structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most popular
criterion has been the marginal likelihood (usually called BDeu for
reasons explained later), but studies~\cite{cosco.uai07,Steck08} have
found this criterion to be very sensitive to hyperparameters and to
yield undesirably complex models for small sample sizes.

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be a potential candidate
for a good criterion, but its exact calculation is likely to be
prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike the fNML,
the qNML gives equal scores to structures that encode the same
independence statements. Like other common model selection criteria,
qNML is also consistent.

>>>>

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information criterion
(AIC)~\cite{Akai73} and the Bayesian information criterion
(BIC)~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for AIC, and BIC appears to
require large sample sizes in order to identify appropriate
structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most
popular criterion has been the Bayesian marginal
likelihood~\cite{Heck95b} and its BDeu variant (see
section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
found this criterion to be very sensitive to hyperparameters and to
yield undesirably complex models for small sample sizes.

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be a potential
candidate for a good criterion, but its exact calculation is likely to
be prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike fNML,
qNML is \textit{score equivalent}, i.e., it gives equal scores to
structures that encode the same independence and dependence
statements. Like other common model selection criteria, qNML is also
consistent.

### 2017-04-01 qnml.tex
finetuning + Teemu's comments
Finetuning

Because the maximum likelihood for the data $D$ decomposes,
we can write
\begin{eqnarray}
  \lefteqn{P_{NML}(D;G)} \nonumber \\
  && = \frac{P(D;\hat\theta(D),G)}
            {\sum_{D_{V_1}}\ldots \sum_{D_{V_C}}\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} \nonumber \\
            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
            {\prod_{c=1}^C\sum_{D_{V^c}}P(D_{V^c};\hat\theta(D_{V^c}),G)}.\nonumber \\
            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
\end{eqnarray}

@@

\textit{
With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
number of different parts of the $i^\text{th}$ partition of $n$,
$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
partition of $n$, one has:
}
$$
a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
$$
For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.

>>>>

Because the maximum likelihood for the data $D$ decomposes,
we can write
\begin{eqnarray}
  \lefteqn{P_{NML}(D;G)} \nonumber \\
  && = \frac{P(D;\hat\theta(D),G)}
            {\sum_{D'_{V_1}}\ldots \sum_{D'_{V_C}}\prod_{c=1}^C P(D'_{V^c};\hat\theta(D'_{V^c}),G)} \nonumber \\
            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
            {\prod_{c=1}^C\sum_{D'_{V^c}}P(D'_{V^c};\hat\theta(D'_{V^c}),G)}.\nonumber \\
            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
\end{eqnarray}

@@

\textit{With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
number of different parts of the $i^\text{th}$ partition of $n$,
$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
partition of $n$, one has:}
$$
a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
$$
For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.
